Relationship Between Common Stream Health Indicators

### 2021-12-09 products/manuscript/Manuscript.Rmd
Manuscript Edits - CC

</br>

Stream health and water quality is typically measured by its physical characteristics (chemistry, surrounding land use area, channelization, etc.); Microbial (fecal indicator), and ecology (diversity of indicator macroinvertebrates). However, these categories are usually analyzed separately to assess stream health. If the impact of urbanization on freshwater bodies is not monolithic, we could be spending a lot of time and effort on monitoring programs that are not effective. Studies have shown conflicting results on what variables like conductivity, turbidity, e. coli cfu (fecal coliform indicator), nutrient (nitrogen and phosphorus) concentration, and biodiversity indicator score actually say about water quality. Therefore, a study was conducted assess how well all these parameters correlated with each other in the context of the Adopt-a-Stream method. Specifically, can you use one of the indicators mentioned (physical, microbial, and ecological) to describe the others to make a statement about overall stream health?

@@

</br>

@@

UOWN has provided a great deal of data dating back to 2003. First, I will need to wrangle missing data values that are a result of inconsistent sampling efforts. This will make the data easier to work with and narrow down a time frame for the data that is consistent and free of sampling bias. Next, I will analyze and compare trends in the change of chemical (conductivity, turbidity,	no3.mgL); ecological (index score based on measurement of stream macroinvertebrate diversity); and microbial indicator data (e.coli CFU). In the context of this study, will will be measure ecological diversity based on the Adopt-A-Stream's method of describing stream health based on diversity of susceptible and tolerant taxa (@GAEPD2015). The idea is that the more urban impact to a stream, the lower the richness of macroinvertebrate taxa sensitive to urban pollution. Generally speaking, increases in chemical properties like conductivity (measure of total dissolved ions), turbidity (measure of suspended solids),	nitrate concentration (no3.mgL) and abundance of microbial indicator taxa of human waste impact (i.e. E. coli cfu) indicate decreased stream health. While a decrease in biological score indicates decrease in stream health. Next, since biological scores are used as a measure of broader ecological health, I will run t-tests on if the other factors significantly effect biological score. Finally, I will test the interaction of all three via multivariate analysis.

</br>

@@

At this point, the main take-aways from data analysis thus far is that correlation of stream health indicators showed that most interactions were non-significant, with the exception of E. coli (cfu) and turbidity. Additionally, the data set is most likely not skewed by influential outliers.

Following correlation analysis during exploration, focus was shifted to just the MIDO dataset. As you can tell, working with many variables in three different data sets can be quite cumbersome to read through. For ease of interpretation, the MIDO data set will serve as the data frame where I test how robust the linear fit models actually are for the data. MIDO is the only data set large enough that when we split the data set into a "test" and "train" data set, I actually have confidence in the resulting model fit analysis. For comparison, MIDO has an n = 88; NORO has an n = 35; and BICO has an n = 23. The only thing we are missing by dropping the other sites is assessing site-by-site comparison, which I am choosing not to focus on anyway.

@@

_Figure 1: Plots of biological score versus conductivity, E. coli cfu, nitrate concentration, and turbidity for the MIDO site._
```{r, echo=FALSE, message=FALSE}
bs_MIDO_location <- here::here("results", "MIDO_bs_lm.png")

@@

</br>

The null linear models for biological score and E. coli cfu (Table 3 and 5) produced a RMSE = 6.63, with a standard deviation of 0.13 and RMSE = 0.52, with a standard deviation of 0.014, respectively. Results from the biological score LASSO machine learning model (Table 4 and 6) indicated that model 47 is the best performing models (RMSE = 6.50; STE = 0.16). However, based on comparative RMSE scores, model 47 does not really perform any better than the null model. This indicates that it is a bad fit to the MIDO data. Results from the E. coli cfu LASSO machine learning model indicated that model 41 is the best performing models (RMSE = 0.50; STE = 0.012). However, based on comparative RMSE scores, model 41 does not really perform any better than the null model. This indicates that it is a bad fit to the MIDO data.

@@

</br>

```{r, echo=FALSE, message=FALSE}
Null_Met_EC_location <- here::here("results", "Null_Met_EC_location.RDS")

>>>>

Stream health and water quality are typically measured by a stream's physical characteristics (chemistry, surrounding land use area, channelization, etc.); microbial impacts(fecal indicator), and ecology (diversity of indicator macroinvertebrates). However, these categories are usually analyzed separately to assess stream health. If the impact of urbanization on freshwater bodies is not monolithic, we could be spending a lot of time and effort on monitoring programs that are not effective. Studies have shown conflicting results on what variables like conductivity, turbidity, e. coli cfu (fecal coliform indicator), nutrient (nitrogen and phosphorus) concentration, and biodiversity indicator score actually say about water quality. Therefore, a study was conducted assess how well all these parameters correlated with each other in the context of the Adopt-a-Stream method. Specifically, can you use one of the indicators mentioned (physical, microbial, and ecological) to describe the others to make a statement about overall stream health?

@@

@@

UOWN has provided a great deal of data dating back to 2003. First, I wrangled missing data values that are a result of inconsistent sampling efforts. This made the data easier to work with and narrowed down a time frame for the data that is consistent and free of sampling bias. Next, I analyzed and compared trends in the change of chemical (conductivity, turbidity,	no3.mgL); ecological (index score based on measurement of stream macroinvertebrate diversity); and microbial indicator data (e.coli CFU). In the context of this study, I measured ecological diversity based on the Adopt-A-Stream's method of describing stream health based on diversity of susceptible and tolerant taxa (@GAEPD2015). The idea is that the more urban impact to a stream, the lower the richness of macroinvertebrate taxa sensitive to urban pollution. Generally speaking, increases in chemical properties like conductivity (measure of total dissolved ions), turbidity (measure of suspended solids),	nitrate concentration (no3.mgL) and abundance of microbial indicator taxa of human waste impact (i.e. E. coli cfu) indicate decreased stream health. In contrast a decrease in biological score indicates decrease in stream health. Next, since biological scores are used as a measure of broader ecological health, I will run t-tests on if the other factors significantly effect biological score. Finally, I will model the interaction of all physical/chemical measurments of stream health on biological score and E. coli cfu via LASSO modeling.

@@

Following correlation analysis during exploration, focus was shifted to just the MIDO dataset. For ease of interpretation, the MIDO data set will serve as the data frame where I test how robust the linear fit models actually are for the data. MIDO is the only data set large enough that when we split the data set into a "test" and "train" data set, I actually have confidence in the resulting model fit analysis. For comparison, MIDO has an n = 88; NORO has an n = 35; and BICO has an n = 23. The only thing we are missing by dropping the other sites is assessing site-by-site comparison, which I am choosing not to focus on anyway.

@@

_Figure 1: Plots of biological score versus conductivity, E. coli cfu, nitrate concentration, and turbidity for the MIDO site._
```{r, echo=FALSE, message=FALSE, out.height="50%"}
bs_MIDO_location <- here::here("results", "MIDO_bs_lm.png")

@@

The null linear models for biological score and E. coli cfu (Table 3 and 5) produced a RMSE = 6.63, with a standard deviation of 0.13 and RMSE = 0.52, with a standard deviation of 0.014, respectively. Results from the biological score LASSO machine learning model (Table 4 and 6; Figure 7) indicated that model 47 is the best performing models (RMSE = 6.50; STE = 0.16). However, based on comparative RMSE scores, model 47 does not really perform any better than the null model. This indicates that it is a bad fit to the MIDO data. Results from the E. coli cfu LASSO machine learning model indicated that model 41 is the best performing models (RMSE = 0.50; STE = 0.012). However, based on comparative RMSE scores, model 41 does not really perform any better than the null model. This indicates that it is a bad fit to the MIDO data.

@@

\newpage
```{r, echo=FALSE, message=FALSE}
Null_Met_EC_location <- here::here("results", "Null_Met_EC_location.RDS")
